Structuring College Choice: The Curious, Mundane Case of College Board Geo-markets

### 2024-07-08 text/old_spring2024/cb_geomarket_lit_review.qmd
initial commit, copying files from messed up repo

---
title: "Structuring College Access: College Board Geo-markets and the Market Segment Model"
author: ["Ozan Jaquette (UCLA)", "Karina Salazar (University of Arizona)"]
bibliography: ./../assets/bib/cb_geomarket_bib.bib
csl: ./../assets/bib/apa.csl
format:
  pdf:
    documentclass: article
    #papersize: letter
    number-sections: true
    number-depth: 2
    #toc: FALSE
    #pandoc_args: !expr rmdfiltr::add_wordcount_filter(rmdfiltr::add_citeproc_filter(args = NULL))
    #keep_tex: true
editor: visual
geometry:
  margin=1in
fontsize: 11pt
header-includes:
  - \usepackage{setspace}
  - \doublespacing
---

@@

# Introduction

hey ho This many words: {{< words-total >}}

COULD START WITH HISTORICAL VIGNETTE OF ZEMSKY PARTNERING WITH COLLEGE BOARD TO CREATE THE MARKET SEGMENT MODEL.

-   quotes on "admissions officers are invariably tellers of stories" and the goal of the market segment model is to quantify these stories.

@@

```{=tex}
\pagenumbering{roman}
\newpage
\pagenumbering{gobble}
\pagenumbering{arabic}
```
# Background: Enrollment Management

@@

The process of recruiting students involves many "make or buy" decisions about whether to perform a given task in-house or outsource it to a third-party vendor. EM consulting firms provide advice and implementation in the areas of marketing, recruiting, pricing and financial aid, and student success. As "creating a class" has becomes complicated and high-stakes [@RN3519], many colleges hire EM consulting firms to develop and/or implement recruiting campaigns. The two largest enrollment consulting firms -- Ruffalo Noel Levitz and EAB -- claim to serve more than 3,000 colleges and universities collectively [@rnlabout; @eabcomm]. In our data collection, attempting to obtain data about student list purchases from all public universities in four states, at least half of these universities outsourced student list purchases to an EM consultancy [@list_biz; @list_empirics].

@@

Scholarship at the nexus of enrollment management and college access can be categorized by which part(s) of the enrollment funnel they focus on. The majority of scholarship focuses on the admissions stage, for example which admissions criteria are used [@RN4874; @RN4993] and which applicants are admitted [@RN4131; @RN3522] \[CITE SOMETHING BY BASTEDO\].

A growing literature analyzes the earlier "recruiting" stages of identifying leads, soliciting inquiries, and soliciting applications. Scholarship on recruiting has conceptualized recruiting behavior as an indicator of college enrollment priorities [@RN4758] and identifies recruiting practices as a mechanism for social reproduction in college access [@RN4324; @RN3519].

Ethnographies by @RN3519 and by @RN4407 analyze the connections between privileged colleges and privileged high schools. @RN4407 shows how guidance counselors at elite private school guidance counselors lobby admissions counselors on behalf of marginal students. @RN3519 finds that college admissions officers visit high schools as a means of maintaining positive relationships with guidance counselors at affluent feeder schools. Consistent with this finding, quantitative case studies of off-campus recruiting visits show that selective private colleges visit affluent, predominantly white schools and communities, with a disproportionate number of visits to private high schools [@RN5014; @RN4330]. Additionally, public research universities often make make more out-of-state recruiting visits than in-state visits, and these out-of-state visits focus on the same set sorts of schools targeted by selective private universities [@RN4759; @RN4758].

@@

We make two observations about the recruiting literature. First, consistent with scholarship on political economy [@RN4517; @cottom2017lower] \[cite Stevens; cottom\] and industrial organization \[cite Winston\], the recruiting literature suggests a hierarchical competition in which different "types" of postsecondary institutions target hierarchically arranged customer niches. For example, for-profit institutions target adults in communities not being targeted by traditional colleges [@cottom2017lower]. Non-selective private non-profits target prospects who might be considering community college or a local state colleges [@RN4324]. Selective private colleges target private high schools, with a particular link between sectarian high schools and colleges of the same faith [@RN5014], but these schools are increasingly targeted by out-of-state public research universities, suggesting growing competition for a coveted market niche. This process of hierarchically organized colleges matching to hierarchical student segments exemplifies the idea that higher education is mechanism of social reproduction [@RN2377; @RN4873]

Second, by focusing on the behaviors of individual colleges, the literature makes the implicit assumption that recruiting is something done by individual colleges. Why then, does recruiting behavior look so similar across colleges of a particular "type"?[^1]

[^1]: New institutional theory offers one explanation [@RN527]. In highly developed organizational fields like higher education, forces of competitive and institutional isomorphism -- including institutional locics -- compel organizations adopt similar practices as their peers. Practices deemed advantageous spread quickly, for example, public research universities targeting nonresident prospects to compensate for declines in state appropriations. These behaviors are conceived as autonomous decisions by individual colleges, albeit structured by institutional logics.<!--WHAT DOMINA (2017) SAID ABOUT LOGICS, P. 322: Like other distinctively modern institutions, contemporary schools are governed by standardized and rationalized institutional logics (Baker 2014, Bowker & Star 2000, Timmermans & Epstein 2010, Meyer 1977). These logics have profound implications for educational category assignment processes, leading educators, policy makers, scholars, and others to debate the universality of selection-->

@RN5012 argues that the process of college access is structured by third-party recruiting products utilized by colleges. @RN5012 deconstructs the College Board *Student Search Service* product. Student list products are "selection devices" [@RN4778] that incorporate search filters (e.g., high school graduating class, state) which enable colleges to select prospective students. Drawing from @RN4786, @RN5012 argue that several search filters (e.g., AP test score, zip code, geodemographic segment) are "racialized inputs," defined as ostensibly race-neutral inputs that are correlated with race because disadvantaged racial groups have historically been excluded from the input. Using a national sample of high school students from the High School Longitudinal Study of 2009 (HSLS), they show that racialized search filters have a strong negative relationship with the selection of Black and Latinx prospects. Using data from actual student lists purchased by public universities, they show that raciaized search filters -- used in conjunction with other search filters -- yield dramatic racial inequality in which prospective students are targeted.

[NEED CLEANER TRANSITION]Two geographic search filters -- geo-market and geodemographic segment -- that @RN5012 observe in lists purchased by public could not recreate using HSLS. Using geographic borders to determine selection for an intervention has been widely studied as a mechanism of structural inequality because of historic and contemporary racial segregation in American communities and schools [e.g., @RN4775; @RN4849; @RN4779]. The geo-market and geodemographic segment search filters of the College Board *Student Search Service* are examples of creating new geographic borders for the purpose of targeting students for educational opportunities. The geo-market filter slices states and metropolitan areas into smaller, "local" markets. For example, CA08 refers to "Alameda Country excluding Oakdland" and a university might purchase prospect contact information by filtering on this geographic region, alongside additional filters. The geodemographic segment filters utilize cluster analysis to allocate each census tract and each high school into different categories based past college enrollment and other factors. This manuscript focuses on geo-markets. The first question we asked ourselves was, "where did these geo-markets come from?" The answer is, market research.

```{=html}
<!--The enrollment management industry emphasizes market research.
Market research is the process of producing information that informs decisions about product design or how to find customers.   -->
```
The enrollment management industry puts great stock in market research. @RN4986 define market research as all "activities intended to increase understanding of a market and consumer behavior in that market" (p. 17). At the most basic level, (in-house) market research is a function of intentional record-keeping. @RN3519 describes the archives of annual reports of admissions activities from a selective liberal arts college. In the 1942 report, "Section 2.5 is titled 'Secondary School Quartile Standings," a measure of where matriculants fell in the rankings of their high school classes" [@RN3519] while Section 3.11 describes the high schools visited. In the chapter entitled "Travel," @RN3519 [p. 69] describes how The College used record-keeping to inform decisions about visits to local high schools in the 2000s: "The College keeps files on hundreds of schools across the country, and around the globe, with which it had contact \ldots The files enable officers to quickly discern"how we've done here lately," \ldots with the number, relative quality, and yield rate of applicants from a particular school."

The College Board played a pivotal role in transforming market research about enrollment from something done by individual universities to something done by third-party professionals. In the late 1970s, College Board provided Carleton College with funding and data for a six-year project that analyzed college choice decisions by households in six metropolitan areas.[^2] This project was the basis for the landmark book *Applying Market Research in College Admissions* [@RN4986], published by College Board, and was the basis for the College Board Admitted Student Questionnaire (ASQ), a revenue-generating product that provides colleges with information about admitted students -- those who enrolled and who did not enroll. <!--Historically, College Board partnered with Members to develop projects that later became revenue-generating products for College Board.    -->

[^2]: In the 1970s, Carleton College became a leader in applying market research to student recruiting, under the leadership of Larry Litten, the then director of institutional research. Litten would later become Associate Director of the Consortium of Financing Higher Education (COFHE).

Similarly, in 1979 The College Board provided Robert Zemsky, a professor of education at UPenn, with data and funding to develop the *Comprehensive Undergraduate Enrollment Planning Project* [@RN4982]. In 1983, The College Board published *The Structure of College Choice* by Robert Zemsky and Penney Oedel, developed the "market segment model" and defined local "geo-markets" based on a comprehensive analysis of SAT score-sending by prospective students to coleges and universities. In 1984/5, the market segment model and geo-markets became the basis for the College Board Enrollment Planning Service (EPS), the market research division of College Board.

This manuscript analyzes the market segment model and geo-markets as a case study in order to investigate market research -- and its commodification into recruiting products -- as a mechanism for social reproduction. At root, third-party market research helps universities identify who are the desirable students, where are they located, which institutions are they considering, and how best to reach them. Today, Ruffalo-Noel Levitz and EAB both produce prodigious amounts of market research, some publicly available as a means of business development \[e.g., CITE RNL\], but most only available to clients. EPS remains the market research arm of College Board, while Encoura Eduventures is the market research arm of ACT. Despite being central to enrollment management for at least 40 years [@RN4986], the role of third-party market research has been ignored by scholarship on college access, which views enrollment management behavior as the autonomous decisions of individual colleges, albeit structured by institutional logics [@RN4758; @RN4873; @RN3519]. We argue that market research structures who colleges recruit, in ways that reinforce existing social stratification.

# Quantification

```{=html}
<!--
CUT GENERAL PARAGRAPH ON REPRODUCTION

Within the sociology of education, @RN2377 describes education as a sieve, allocating people to opportunities in socially stratified, capitalist societies. This role highlights "Max Weber's core insight that education has a dual character" [@RN2377, p. 129], on one hand, serving social mobility by redistributing access to privileged positions and, on the other, serving social reproduction by granting privileged families disproportionate access to the educational institutions that confer pathways to privileged positions. Similarly, @RN4873 describe schools and colleges as "sorting machines" that incorporate external categories (e.g., race, class), create categories (e.g., instructional tracks), and allocate students to categories. Initial category allocation affects subsequent allocation, such that school assignments and instructional track assignments affect college access, which affects employment assignments. "In evaluating schools as inequality regimes" @RN4873 \[p. 322\] argue, "the relevant question is not whether schools produce inequality; rather, the relevant questions are, when and where do schools produce inequality, \[and\] how do they do so." As @RN2377 show, whereas early research on social stratification sought to identify patterns of inequality [e.g., @RN4873; @RN290; @RN1268], the sociology of education is now more concerned with identifying the mechanisms underlying these patterns. 
-->
```
We situate our study amidst scholarship on within scholarship on quantification as a mechanism that reproduces inequality. @RN5007 [p. 402] define quantification as "the production and communication of numbers" (p. 402) and introduce reacivity, discipline, and authority as three interrelated themes that describe the effects of quantification. Reactivity is the idea that salient quantitative measures cause people and organizations to change their behavior. Discipline is the idea that quantification causes actors to react in a particular way. Authority is the idea that quantification changes decision-making power, often weakening the discretion of local decision-makers.

Scholarship about U.S. News & World Report (USNWR) Law School rankings demonstrate the reactivity, discipline, and authority effects of quantification [@RN5003; @RN2356; @RN2376; @RN2375]. Developed as a means of informing prospective students about quality, rankings affected which law schools applicants applied to, the intended effect. However, rankings also affected hiring decisions of firms because clients evaluated firms based on the prestige of law schools attended by firm lawyers. (clients evaluated firms based on the prestige of the law schools that firm lawyers attended) and

USNWR developed the rankings as a means to inform prospective students about law school quality. However, rankings affected

US News developed law schools rankings as a means of informing prospective students about law school quality. In turn, rankings affected how prospective students conceive law school quality and which schools they applied to. Rankings also affected the cognition and behavior of other actors. Looking downstream, entities that received law school students as an input -- for example, law firms -- used rankings as a hiring criterion because their clients evaluated the organization based on the the prestige of the schools their firm's lawyers attended.

Drawing from Foucault \[CITE DISCIPLINE AND PUNISH\], discipline is the idea that surveilance -- attributes of a law school being quantified and aggregated into a public-facing metric -- compels people and organizations to react in a particular way. For example, the reputational assessment of peer leaders as an important component of most US News rankings. Bastedo and Bowman showed that once (undergraduate) USNWR rankings became popular, the reputational assessment of peers primarily became a function of previous USNWR rankings. Once law schools realized that "important groups of constituents --- students, faculty, trustees, employers, other media --- were using rankings to make decisions that had large consequences for schools \ldots schools felt pressured to take them seriously" \[QUOTE FROM ESPELAND AND STEVENS\]. For example, applicants with characteristics valued by the rankings system (e.g., LSAT scores) became more important for decisions about admissions and merit aid \[CITE\].

The theme of authority is that quantification destabilizes and weakens the authority of local decision-makers. For example, law school rankings weakened the authority of law schools to make independent decisions about which applicants to admit, ideal class size, and curricular offerings \[CITE THE BOOK\]. In a process akin to sports analytics in Moneyball \[CITE MICHAEL LEWIS\], we suggest that market research replaces the wisdom and authority of local admissions counselors.

The analysis of UK League Tables by @RN5005 show how quantification can be a mechanism of social reproduction. In 1992, as part of a the broader *New Public Management* effort to instill market competition amongst public providers [@RN161], the UK government started publishing school "league tables" that ranked schools based on student performance on the high-stakes General Certificate of Secondary Education (GCSE) exams that students take at age 16. @RN5005 argue that quantifying the quality of schools entrenches class-based inequalities between schools particularly when the measure of quality -- standardized test scores -- is substantially a function of the class composition of schools. School league tables discipline households and schools to conceive of school quality in terms of test scores and they facilitate making evaluative comparisons between schools that are geographically close or distant. Compared to working-clas parents, parents with professional/managerial occupations are more aware of school league tables and have financial resources necessary to move to neighborhoods near a "good" school.

Using school-level data and data on census data on occupational class measured at the local authority level (similar to a U.S. county), @RN5005 find that, following the adoption of league tables, local areas with higher performing schools experience growth in the share of managerial/professional occupational classes and a decline in the share of working class families. Additionally, using longitudinal survey data on 1% of the population of England and Wales and a diference-in-difference estimator, the authors find that managerial/professional households are more likely to move to local areas with higher ranked schools following the introduction of school league tables. Overall, @RN5005 show that quantifying which schools are high-performing contributes to social reproduction because managerial/professional households respond to consumer-facing metrics by moving to more expensive neighborhoods, near higher performing schools. Schools that performed well on initial league tables tended to be schools with a high share of professional/managerial families. Therefore, quantification takes a snapshot of existing social stratification and that snapshot becomes a power unto itself, amplifying the effect of previous strctural inequality on subsequent structural inequality.

In the broader quantification literature, critical data studies [e.g., @RN4975; @RN4849; @RN4772] and sociological analyses of technology and inequality [e.g., @RN4778; @RN4786; @RN4877; @RN4810] introduce concepts salient to the analysis of market research. Our focus is on "selection devices," which utilize some combination of inputs and discretion to categorize places, people, and organizations for the purpose of allocating opportunities or interventions [@RN4778; @RN5012].

A multi-disciplinary literature examines structural inequality produced by predictive analytics that capture correlational relationships [@RN4975]. Predictive analytics utilize actuarial methods, which were pioneered by the insurance industry and proceed in two steps [@RN4835; @RN4778]. First, apply statistical techniques to previous cases (training data) in order to identify factors positively and negatively associated with an outcome of interest. Second, apply these results to future cases in order to make predictions and to assign levels of risk to each case. Importantly, these models predict outcomes based on correlations without requiring knowledge about causal relationships. @RN4975 provides the example of @RN4989, who develop a method to predict sensitive personal attributes (e.g., gender, political party) based on Facebook Likes.

The use of predictive analytics reproduces structural inequality. The correlations observed during the training data stage are a snapshot of relationships between variables at a particular point of time. Although these correlations may b e a function of enduring structural inequality, the underlying causes are not considered by applications of predictive models.[^3]. This algorithm recommends high purchase volume in zip codes that sent large number of students in previous years. @RN4794 \[p. 224\] state that "predicting the future on the basis of the past threatens to reify and reproduce existing inequalities. @RN4851 refers to this phenomenon as the"ratchet effect," whereby disproportionately targeted/excluded populations are predicted to have a higher risk of an outcome, which amplifies disproportionate targeting/exclusion.[HMM. PERNICIOUS FEEDBACK LOOPS WOULD SEEM TO FIT HERE]

[^3]: For example, the College Board *Student Search Service* allows colleges to filter student list purchases based on zip code, amongst other filters. The consulting firm Ruffalo Noel Levitz offers an algorithm that recommends how many prospect profiles a college should purchase from each zip code, based on historic enrollment data from the college [@jmu_rnl]

@RN4849 shows how predictive analytics create "weapons of math destruction" (WMD), which are defined as algorithms that have three characteristics: opacity, meaning that the algorithm is proprietary or unknown to the people it affects; scale, meaning that the algorithm can affect large numbers of people; and damage, meaning that the algorithm can reinforce past inequality. An exemplar WMD is the XXX model used to predict a prisoner's chances of re-arrest. How the algorithm uses inputs to predict the outcome is unclear (opacity).[^4] The algorithm is used by 24 states (scale). Because the algorithm uses zip code as an input, people who live in highly policed neighborhoods have a higher predicted probability of being arrested, which leads to more policing in those neighborhoods, which perpetuates racialized inequality in arrests. @RN5006 observe that such "pernicious feedback loops" [@RN4849] are similar to the "ratchet effect" [@RN4851] and are extreme versions of "reactivity" from @RN5003.

[^4]: By contrast, U.S. law school rankings and UK leage tables are not WMDs because these algorithms are transparent [@RN5006].

Due to data limitations, weapons of math destruction (WMD) -- and predictive analytics more broadly -- often utilize "proxy" variables, which are variables that are correlated to a variable of interest [@RN4849]. FICO credit scores are based solely on individual repayment history and do not utilize proxies. By contrast, "e-scores" utilize proxy variables to identify "people like you" and then predict your buying behavior based on the past buying behavior of these others. @RN4849 [p. 146] states that "the modelers for e-scores have to make do with trying to answer the question 'How have people like you behaved in the past?' when ideally they would ask, 'How have you behaved in the past?'" We observe similar behavior in market research about college access, when enrollment managers use 'which colleges did students near you consider' as a proxy for, 'which colleges would you consider?'

@RN4975 argues that ideas about homophily often underlie the use of correlation to predict future behavior. Homophily means that actors who share common characteristics are likely to form connections with one another. The ubiquitous colloquialism is "birds of a feather flock together." Homophily has become a core assumption for network science [@mcpherson2001birds], in which actors (e.g., high school students) are linked to one another through direct and indirect network ties (e.g., submit SAT scores to the same college). Because network science often draws from rational choice theory, it assumes that homophily is the result of voluntary action by individuals, which "erases historical contingencies, institutional discrimination, and economic realities" [@RN4975, p. 95] that underlie behavior consistent with homophily. @RN4975 problematizes the idea that homophily is a naturally occurring phenomenon. In commercial social networks, such as Facebook, Twitter/X, and TikTok, homophily is programmed into the algorithms that create connections between users. Thus, "echo chambers are not unfortunate errors, but deliberate goals" because "homophily is used to create agitated clusters of individuals whose angry similarity and overwhelming attrication to their common object of hatred both repel them from one another and glue them together" \[p. 82\].

>>>>

---
title: "Structuring College Choice: The Curious, Mundane Case of College Board Geo-markets"
author: ["Ozan Jaquette (UCLA)", "Karina Salazar (University of Arizona)"]
bibliography: ./../assets/bib/cb_geomarket_bib.bib
csl: ./../assets/bib/apa.csl
format:
  pdf:
    documentclass: article
    #papersize: letter
    number-sections: true
    number-depth: 2
    #toc: FALSE
    #pandoc_args: !expr rmdfiltr::add_wordcount_filter(rmdfiltr::add_citeproc_filter(args = NULL))
    #keep_tex: true
editor: visual
geometry:
  margin=1in
fontsize: 12pt
header-includes:
  - \usepackage{setspace}
  - \doublespacing
---

@@

```{=tex}
\pagenumbering{roman}
\newpage
\pagenumbering{gobble}
\pagenumbering{arabic}
```
# Introduction

@@

# Background: Enrollment Management and the Enrollment Funnel

@@

@@

Scholarship at the nexus of enrollment management and college access can be categorized by which part(s) of the enrollment funnel they focus on. The majority of scholarship focuses on the admissions stage, for example which admissions criteria are used [@RN4874; @RN4993] and which applicants are admitted in practice [@RN4131; @RN3522] \[CITE SOMETHING BY BASTEDO\].

A growing literature analyzes the earlier "recruiting" stages of identifying leads, soliciting inquiries, and soliciting applications. Scholarship on recruiting has conceptualized recruiting behavior as an indicator of college enrollment priorities [@RN4758] and identifies recruiting practices as a mechanism for social reproduction in college access.

Ethnographies by @RN3519 and by @RN4407 analyze the connections between privileged colleges and privileged high schools. @RN4407 shows how guidance counselors at elite private school guidance counselors lobby admissions counselors on behalf of marginal students. @RN3519 finds that college admissions officers visit high schools as a means of maintaining positive relationships with guidance counselors at affluent feeder schools. Consistent with this finding, quantitative case studies of off-campus recruiting visits show that selective private colleges visit affluent, predominantly white schools and communities, with a disproportionate number of visits to private high schools [@RN4733; @RN4330]. Additionally, public research universities often make make more out-of-state recruiting visits than in-state visits, and these out-of-state visits focus on the same set sorts of schools targeted by selective private universities [@RN4759; @RN4758].

@@

We make two observations about the recruiting literature. First, looking across studies, the recruiting literature suggests a hierarchical competition in which different "types" of postsecondary institutions target different customer niches. For example, for-profit institutions target adults in communities not being targeted by traditional colleges [@cottom2017lower]. Non-selective private non-profits target prospects who might be considering community college or a local state colleges [@RN4324]. Selective private colleges target private high schools, with a particular link between sectarian high schools and colleges of the same faith [@RN4733], but these schools are increasingly targeted by out-of-state public research universities, suggesting growing competition for a coveted market niche.

Second, the recruiting literature -- and scholarship on enrollment management more broadly -- assumes that recruiting is a function of individual organizational behavior \[alternatively, "assumes that recruiting is something done by individual colleges"\]. Why then, does recruiting behavior look so similar across colleges of a particular "type"? Institutional theory offers a compelling explanation [@RN527]. In highly developed organizational fields like higher education, forces of competitive and institutional isomorphism -- including institutional locics -- compel organizations adopt similar practices as their peers. Practices deemed advantageous spread quickly, for example, public research universities targeting nonresident prospects to compensate for declines in state appropriations. These behaviors are conceived as autonomous decisions by individual colleges, albeit structured by institutional logics. <!--WHAT DOMINA (2017) SAID ABOUT LOGICS, P. 322: Like other distinctively modern institutions, contemporary schools are governed by standardized and rationalized institutional logics (Baker 2014, Bowker & Star 2000, Timmermans & Epstein 2010, Meyer 1977). These logics have profound implications for educational category assignment processes, leading educators, policy makers, scholars, and others to debate the universality of selection-->

Whereas most scholarship assumes that recruiting is something done by individual colleges, @RN4985 argues that structural inequality is embedded in third-party recruiting products utilized by colleges. @RN4985 deconstructs the College Board *Student Search Service* product. Student list products are "selection devices" [@RN4778] that incorporate search filters (e.g., high school graduating class, state) which enable colleges to select prospective students. Drawing from @RN4786, @RN4985 argue that several search filters (e.g., AP test score, zip code, geodemographic segment) are "racialized inputs," defined as ostensibly race-neutral inputs that are correlated with race because disadvangaged racial groups have historically been excluded from the input. Using a national sample of high school students, they show that racialized search filters have a strong negative relationship with the selection of Black and Latinx prospects. Using data from actual student lists purchased by public universities, they show that raciaized search filters -- used in conjunction with other search filters -- yield dramatic racial inequality in which prospective students are targeted.

# OLD SHIT

We situate this scholarship broadly within the literature on social reproduction and narrowly within the literature on inequality in colege access, with a particuar focus on enrollment management behavior.

@RN2377 describes education as a sieve, allocating people to opportunities in socially stratified, capitalist societies. This role highlights "Max Weber's core insight that education has a dual character" [\@ @RN2377, p. 129], on one hand, serving social mobility by redistributing access to privileged positions and, on the other, serving social reproduction by granting privileged families disproportionate access to the educational institutions that confer pathways to privileged positions. Similarly, @RN4873 describe schools and colleges as "sorting machines" that incorporate external categories (e.g., race, class), create categories (e.g., instructional tracks), and allocate students to categories. Initial category allocation affects subsequent allocation, such that school assignments and instructional track assignments affect college access, which affects employment assignments. "In evaluating schools as inequality regimes" @RN4873 \[p. 322\] argue, "the relevant question is not whether schools produce inequality; rather, the relevant questions are, when and where do schools produce inequality, \[and\] how do they do so." As @RN2377 show, whereas early research on social stratification sought to identify patterns of inequality [e.g., @RN4873; @RN290; @RN1268], the sociology of education is now more concerned with identifying the mechanisms underlying these patterns. <!--  [NOTE: THIS SETS UP IDENTIFYING MARKET RESEARCH AS AN IMPORTANT **NEW** MECHANISM OF SOCIAL REPRODUCTION THAT SOCIOLOGY OF EDUCATION SHOULD BE STUDYING!!!] -->

Given the substantial social and economic benefits of graduating from college \[CITE\], and the greater benefits of graduating from a selective college \[CITE CHETTY AND STUFF\], sociology has paid particular attention to social stratification in the transition from high school to college. A significant strand of research focuses on cultural capital as a broad mechanism contributing to social reproduction in access to college, particularly selective colleges [@RN1228; @RN1241]. @RN4991 observes that "even as even as colleges' definitions of merit have evolved, from the post-World War II growth of standardized testing,\ldots to a renewed focus on holistic review," affluent, college-educated families have maintained a competitive advantage college access \[CITE\]. These families are socialized to place value on selective colleges, have financial and cultural resources to devote to this competition, and are concentrated in communities and schools that orient student achievement and activities towards the admissions criteria of selective institutions \[CITE\]. <!--  CAN YOU SAY THIS MORE EFFICIENTLY; LIKE ON THE STUDENT DEMAND SIDE, SOCIOLOGY OF EDUCATION HAS FOCUSED ON CULTURAL CAPITAL, WHEREBY PRIVILEGED FAMILIES EMBODY THE CHARACTERISTICS OF DESIRABLE APPLICANTS; AND ON SUPPLY SIDE COLLEGES REALLY WANT THESE APPLICANTS. -->

On the supply-side of college access, an interdisciplinary literature focuses on enrollment management. The term "enrollment management" (EM) refers to a profession, an administrative structure, and an industry. As a profession, EM integrates techniques from marketing and economics in order to "influence the characteristics and the size of enrolled student bodies" [@RN2771, p. xiv]. As an administrative structure, the EM office typically controls the activities of admissions, financial aid, and recruiting [@RN2406]. The EM industry consists of university personnel (e.g., admissions counselors, VP for enrollment management), professional associations (e.g., National Association for College Admission Counseling), and third-party servicers in the consulting, market research, and technology spaces (e.g., College Board, ACT, EAB, Huron, Slate, PowerSchool).

Scholarship at the nexus of enrollment management and college access can be categorized by which part(s) of the enrollment funnel they focus on. The majority of scholarship focuses on the admissions stage, for example which admissions criteria are used [@RN4874; @RN4993] and which applicants are admitted in practice [@RN4131; @RN3522] \[CITE SOMETHING BY BASTEDO\].

A growing literature analyzes the earlier "recruiting" stages of identifying leads, soliciting inquiries, and soliciting applications. Scholarship on recruiting has conceptualized recruiting behavior as an indicator of college enrollment priorities [@RN4758] and identifies recruiting practices as a mechanism for social reproduction in college access.

Ethnographies by @RN3519 and by @RN4407 analyze the connections between privileged colleges and privileged high schools. @RN4407 shows how guidance counselors at elite private school guidance counselors lobby admissions counselors on behalf of marginal students. @RN3519 finds that college admissions officers visit high schools as a means of maintaining positive relationships with guidance counselors at affluent feeder schools. Consistent with this finding, quantitative case studies of off-campus recruiting visits show that selective private colleges visit affluent, predominantly white schools and communities, with a disproportionate number of visits to private high schools [@RN4733; @RN4330]. Additionally, public research universities often make make more out-of-state recruiting visits than in-state visits, and these out-of-state visits focus on the same set sorts of schools targeted by selective private universities [@RN4759; @RN4758].

We make two observations about the recruiting literature. First, looking across studies, the recruiting literature suggests a hierarchical competition in which different "types" of postsecondary institutions target different customer niches. For example, for-profit institutions target adults in communities not being targeted by traditional colleges [@cottom2017lower]. Non-selective private non-profits target prospects who might be considering community college or a local state colleges [@RN4324]. Selective private colleges target private high schools, with a particular link between sectarian high schools and colleges of the same faith [@RN4733], but these schools are increasingly targeted by out-of-state public research universities, suggesting growing competition for a coveted market niche.

Second, the recruiting literature -- and scholarship on enrollment management more broadly -- assumes that recruiting is a function of individual organizational behavior \[alternatively, "assumes that recruiting is something done by individual colleges"\]. Why then, does recruiting behavior look so similar across colleges of a particular "type"? Institutional theory offers a compelling explanation [@RN527]. In highly developed organizational fields like higher education, forces of competitive and institutional isomorphism -- including institutional locics -- compel organizations adopt similar practices as their peers. Practices deemed advantageous spread quickly, for example, public research universities targeting nonresident prospects to compensate for declines in state appropriations. These behaviors are conceived as autonomous decisions by individual colleges, albeit structured by institutional logics. <!--WHAT DOMINA (2017) SAID ABOUT LOGICS, P. 322: Like other distinctively modern institutions, contemporary schools are governed by standardized and rationalized institutional logics (Baker 2014, Bowker & Star 2000, Timmermans & Epstein 2010, Meyer 1977). These logics have profound implications for educational category assignment processes, leading educators, policy makers, scholars, and others to debate the universality of selection-->

Whereas most scholarship assumes that recruiting is something done by individual colleges, @RN4985 argues that structural inequality is embedded in third-party recruiting products utilized by colleges. @RN4985 deconstructs the College Board *Student Search Service* product. Student list products are "selection devices" [@RN4778] that incorporate search filters (e.g., high school graduating class, state) which enable colleges to select prospective students. Drawing from @RN4786, @RN4985 argue that several search filters (e.g., AP test score, zip code, geodemographic segment) are "racialized inputs," defined as ostensibly race-neutral inputs that are correlated with race because disadvangaged racial groups have historically been excluded from the input. Using a national sample of high school students, they show that racialized search filters have a strong negative relationship with the selection of Black and Latinx prospects. Using data from actual student lists purchased by public universities, they show that raciaized search filters -- used in conjunction with other search filters -- yield dramatic racial inequality in which prospective students are targeted.

This manuscript builds on @RN4985 by conceptualizing market research as a mechanism of social reproduction in education. Market research is "the process of gathering, analyzing, and interpreting information about a market, including its consumers, competitors, and the overall industry" \[CHAT GPT, REPLACE\]. Market research is oriented to the goals producers/sellers and is the foundation for marketing efforts and products that target desired prospects. This manuscript shows how College Board "geo-markets" emerged from market research by @RN4982. @RN4982 used SAT score-sending behavior from 19XX to develop the "market segment model," which shows how student demand for different kinds of postsecondary institutions varies across different geographies. Geo-markets carve large metropolitan areas into smaller "local" markets. For each geo-market, the market segment model categorizes students by the kind of postsecondary institutions they are likely to be interested in and calculates institution-specific measures of student demand. In turn, the market segment model is the basis for the College Board Enrollment Planning Service (EPS), which helps colleges plan and implement recruiting strategy.

Empirically, we examine how socioeconomic and racial composition differs across geomarkets. We conceptualize the market segment model as a snapshot of social stratification in student demand for higher education. Additionally, because of segregation and structural racism in income/wealth, we expect local geomarkets to be correlated with race. Later, College Board geo-markets become a filter on the Student Search Service product that enables colleges to purchase the contact information of prospective students. Using simulations, we show how the how the racial and socioeconomic composition of student list purchases change as different geomarkets are selected. We also examine actual student list purchases that filter on geomarket.

This research makes a broad contribution to the sociology of education by conceptualizing market research as a mechanism of inequality. Market research helps producers identify and target potential customers by taking snapshots of existing social stratification in who are the desirable prospective customers and where do they live. These snapshots are a function of historic and contemporary structural inequalities. Market research uses these snapshots to make recommendations about which prospective customers to recruit and to create marketing products. These recommendations an products are then used by producers. Because the snapshots are reflections of historic structural inequality, using these snapshots to identify future customers is likely to amplify historic structural inequality.\[discussion section identifies areas of future research\]

NEXT SUBSECTION IS MARKET RESEARCH AS A MECHANISM OF INEQUALITY IN SOCIOLOGY WRIT LARGE

-   poon, FICO scores
-   geodemography
    -   from EEPA article: using geography as a predictive input was pioneered by geodemography, a branch of market research that estimates the behavior of consumers based on where they live (Burrows and Gane, 2006)
-   fourcade and healey
    -   classification situations
-   key theme is third-party market research and related products that sorts customers for the benefit of merchants.

This study extends @RN4985 in two ways. First, critical scholarship on data science finds that geography is the most common racialized input, owing to centuries of residential segregation. College Board student list product enables colleges to filter based on known geographic units (state, zip code, MSA). But it also creates new academic units ... bla bla

second, PARTX: Scholars have not analyzed the ways that market research is built into products that recruit students In other words, market research identifies student demand for different kinds of institutions This market research is then incorporated into recruiting projects We argue that this phenomenon may be an important mechanism by which historic racial and socioeconomic inequality in student demand is utilized to inform the subsequent recruiting behavior of universities.

We Overall, scholarship from sociology tends to assume that recruiting -- and, in turn, structural inequality in recruiting -- is a function of individual college behavior.

TIFFANY HUANG SAYS THIS IS ABOUT REPRODUCTION; MECHANISMS OF HOW REPRODUCTION OF INEQUALITY HAPPENS; THIS FEELS RIGHT; Tiffany and she helped me realize that the literature I am contributing to (broadly) is social reproduction. more specifically, how (third-party) supply-side actors take snapshots of existing social stratification and build those into the supply-side architecture.

But another mechanism of reproduction is market research and marketing. Market research takes a picture of "student demand" one that is a function of class and racial inequality and then uses that picture to create river beds of where colleges should allow recruiting to flow.
